Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-12-22 parts/sa.tex
fixes and suggestions of Morgan
begin transformation PH to AAN

The aim of this section is to define the problem of reachability in a PH,
and propose an under-approximation allowing to efficiently solve it.
The static analysis presented here is inspired from~\cite{PMR12-MSCS}.

We consider in this section a PH model $\PH = (\PHs; \PHl; \PHa^{\langle 2 \rangle})$ with $2$ classes of priorities
that respects all Conditions (\ref{cr:bounded} \& \ref{cr:compcs}) of \pref{ssec:hypothesis}.

@@

\modMF{
The main idea behind the static analysis presented in this section
is to abstract the dynamics of a Process Hitting model
by a more general and easier to analyse dynamics.
For this, we focus on the notion of objective:
an objective is a couple of processes (\eg $\PHobj{a_i}{a_j}$) whose reachability
is required to ensure the global reachability property considered;
in other words, it is required that a set of actions hitting $a$ exists so that,
starting from a state containing $a_i$,
it is possible to play these actions (possibly intertwined with other actions)
and reach a state containing $a_j$.
Given the particular form of the actions in Process Hitting,
the solving of such an objective raises new objectives in the general case,
because each action can be triggered by (at most) one process from a different sort.
Indeed, each of these processes also have to be reached before the required actions are played.
However, the order of the considered actions is abstracted,
and so is the order of the related objectives,
thus resulting in an over-approximation of the requirements.
}

@@

The reachability of a process $a_j$ of a given sort $a$ from another process $a_i$ is called an objective and is denoted $\PHobjp{a}{i}{j}$ (\pref{def:obj}).
\begin{definition}[Objective ($\Obj$)]
\label{def:obj}
  If $a \in \components$, the reachability of a process $a_j$ from a process $a_i$ is called an \emph{objective}, noted $\PHobj{a_i}{a_j}$.
  The set of all objectives is called $\Obj \DEF \{ \PHobj{a_i}{a_j} \mid a \in \components \wedge (a_i, a_j) \in \PHl_a \times \PHl_a \}$.
  For an objective $P = \PHobj{a_i}{a_j} \in \Obj$, we define: $\PHsort(P) \DEF
  a$, $\PHtarget(P)\DEF a_i$, $\PHbounce(P)\DEF a_j$,
  and $P$ is said \emph{trivial} if $a_i = a_j$.

  We define an \emph{objective sequence} as a sequence of objectives in which each objective target must be equal to the previous objective bounce of the same sort, if it exists.
  The set of all objective sequences is denoted by $\OS$.
\end{definition}

A context (\pref{def:context}) extends the notion of state to a set of possible initial states:
\modMF{
to each sort in the model, a context maps a set of processes in this sort.
}
We also extend the override operator to contexts (\pref{def:ctxcap}).
\begin{definition}[Context ($\Ctx$)]
\label{def:context}
  A \emph{context} $\ctx$ associates to each sort in $\PHs$ a non-empty subset of its processes:
  $\forall a \in \PHs, \PHget{\ctx}{a} \subseteq \PHl_a \wedge \PHget{\ctx}{a} \neq \emptyset$.
  $\Ctx$ is the set of all contexts.
\end{definition}
%
\begin{definition}[$\Cap: \Ctx \times \powerset(\PHproc) \rightarrow \Ctx$]
\label{def:ctxcap}
  For any $\ctx\in\Ctx$ and set of processes $ps \in \powerset(\PHproc)$,
  the override of $\ctx$ by $ps$ is noted $\ctx \Cap ps$ and is defined by:
  \[ \forall a \in \PHs, \PHget{(\ctx \Cap ps)}{a} \DEF
  \begin{cases}
    \{ p \in ps \mid \PHsort(p)=a \} & \text{if } \exists p \in ps, \PHsort(p)=a,\\
    \PHget{\ctx}{a} & \text{else.}
  \end{cases}
  \]
\end{definition}
\noindent
For a given context $\ctx$, we note $a_i \in \ctx$ if and only if $a_i \in \PHget{\ctx}{a}$,
and for all $ps \in \powerset(\PHproc)$ or $ps \in \PHl$, $ps \subseteq \ctx \EQDEF \forall a_i \in ps, a_i \in \ctx$.
A sequence of actions $\delta$ is \emph{playable} in a context $\ctx$ if and only if 
$\exists s \subseteq \ctx, \delta \in \Sce(s)$.
We denote then: $\delta \in \Sce(\ctx)$,
and the play of $\delta$ in $\ctx$ is $\ctx \PHplay \delta = \ctx \Cap \ceil(\delta)$,
where $\ceil(\delta)$ is the set containing the last process in the sequence $\delta$ (hitter or bounce) of every sort mentioned in $\delta$.

Finally, a bounce sequence on a sort $a$ (\pref{def:bs}) is a sequence of actions hitting $a$
in which the bounce process of each action is the hitter process of the following action.
Bounce sequences are used to find local solutions to a given objective:
\modMF{
indeed, the bounce sequences related to an objective $\PHobj{a_i}{a_j}$
only take depend on the bounces of the actions hitting the sort $a$.
}%
A bounce sequence on $a$ can be abstracted into the minimal sets of its hitters that are not in sort $a$ (\pref{def:aBS}).
This abstraction is later used to propagate an objective on a given sort
by creating new objectives on other sorts.
In the following, we denote: $\Sol = \powerset(\PHproc)$.
\begin{definition}[Bounce sequence ($\BS$)]
\label{def:bs}
  A \emph{bounce sequence} $\zeta$ is a sequence of actions so that $\forall n \in \indexes{\zeta}, n < |\zeta|, \PHbounce(\zeta_{n}) = \PHtarget(\zeta_{n+1})$.
  $\BS$ denotes the set of all bounce sequences, and
  $\BS(P)$ denotes the set of bounce sequences \emph{solving} an objective $P$:
  \[
    \BS(\PHobj{a_i}{a_j}) \DEF \{ \zeta \in \BS \mid \PHtarget(\zeta_1)= a_i \wedge \PHbounce(\zeta_{|\zeta|}) = a_j \} \enspace.
  \]
  $\BS(\obj{a_i}{a_j}) = \emptyset$ if there is no way to reach $a_j$ from $a_i$
  and $\emptyseq \in \BS(\obj{a_i}{a_i})$
  \modMF{
  for a trivial objective $\PHobj{a_i}{a_i}$.
  }
\end{definition}
%
\begin{definition}[$\aBS:\Obj \rightarrow \powerset(\Sol)$]
\label{def:aBS}
  The \emph{abstractions of bounce sequences} of an objective $P$, denoted by the set $\aBS(P)$, are the sets of hitters of bounce sequences solving $P$:
  \[
    \aBS(P) \DEF \{ \abstr{\zeta} \in \Sol \mid \zeta \in \BS(P), \nexists \zeta' \in \BS(P), \abstr{\zeta'} \subsetneq \abstr{\zeta} \} \enspace,
  \]
  \modMF{
  where $\forall P \in \Obj, \forall \zeta \in \BS(P)$:
  \[
    \abstr{\zeta} \DEF \{ \PHhitter(\zeta_n) \mid
    n \in \indexes{\zeta} \wedge \PHsort(\PHhitter(\zeta_n)) \neq \PHsort(P) \}
  \]
  }%
\end{definition}

@@

\noindent
The notation $\delta_{a..b}$ denotes the subsequence of $\delta$
between indexes $a$ and $b$, as defined on page \pageref{notations}.

@@

For any objective $P$ and context $\ctx$, \pref{def:maxCont} gives the set of processes of sort $\PHsort(P)$ that are required to solve $P$ in $\ctx$, denoted by $\gCont_\ctx(\PHsort(P), P)$.
\begin{definition}[$\gCont_\ctx : \Sigma \times \Obj \rightarrow \powerset(\PHproc)$]
  \label{def:maxCont}
  \begin{align*}
    \gCont_\ctx(a,P) \DEF
    \{ p \in \PHproc &\mid \exists ps \in \aBS(P), \exists b_i \in ps, b = a \wedge p = b_i \\
      & \vee b \neq a \wedge p \in \gCont_\ctx(a, \PHobj{b_j}{b_i}) \wedge b_j \in \PHget{\ctx}{b} \}
    \enspace.
  \end{align*}
\end{definition}

\modMF{
A graph of local causality $\myB = (\cwV, \cwE)$
is a graph where $\cwV \subseteq \Proc \cup \Obj \cup \Sol$ is the set of vertices
and $\cwE \subseteq \cwV \times \cwV$ is the set of oriented edges.
Such a graph aims a describing the requirements of a given reachability
problem, that is, the reachability of $\w$ from the initial context $\ctx$.
Thus, a node in $\PHproc$ represents a process required to play an action,
a node in $\Obj$ is an objective to reach a given process
and a node in $\Sol$ is a set of hitter processes required for the solving
of an objective.
}

\modMF{
The edges of such a graph allow to draw links between these requirements.
Thus, an objective $P \in \Obj$ is solvable if
at least one related bounce sequence in $\BS(P)$ can be played;
therefore, we can consider all minimal sets of hitters of these bounce sequences
(that is, the sets of $\aBS(P) \in \Sol$, see \pref{def:aBS})
as requirements to solve this objective (\pref{eq:ESol1}).
Such a set can then naturally be split into as many process nodes,
making each of them a new requirement (\pref{eq:ESol2}).
If $a \in \components$, the reachability of a required process $a_i$
is approximated by the ability to solve all objectives of the form
$\PHobjp{a}{j}{i} \in \Obj$ for all $a_j$ in the initial context (\pref{eq:EReq});
if $a \in \cs$, the reachability of $a_i$ is simply solved by the set of processes $\csState(a_i)$ (see \pref{def:csState}) that it represents (\pref{eq:EPrio}).
Furthermore, we note that the solving of an objective $P$ may require
a process of $\PHsort(P)$,
\ie $\gCont(\PHsort(P), P) \neq \emptyset$ (\pref{def:maxCont});
in this case, $P$ is re-targeted (\pref{eq:ECont}).
Finally, \pref{eq:Vw} forces the graph to contain all the initial objectives,
and \pref{eq:VE} ensures the consistency of $\cwV$.
}

\modMF{
However, this graph $\myB$ alone is not sufficient to check the reachability of
an objective sequence $\w$ from an initial context $\ctx$;
indeed, the reachability of each intermediate required process
is only checked starting from the context $\ctx$,
although this context may be modified by other objectives.
Therefore, such a reachability requires the building of the
\emph{saturated graph of local causality} (\pref{def:glc})
which is obtained by iteratively saturating the initial context with all
processes referenced in the graph,
until reaching a fixed point.
Such a fixed point always exists because the size of the context
is limited by the size of the initial PH model.
The set of processes contained in a graph of local causality
is given by:
\begin{align*}
  \allprocs(\cwV, \cwE) = (\cwV \cap \Proc) &\cup \{ \PHtarget(P) \mid P \in \cwV \cap \Obj \}\\
  & \cup \{ P \neq \w \Rightarrow \PHbounce(P) \mid P \in \cwV \cap \Obj \}
\end{align*}
where, if the objective sequence $\w$ consists of only one element,
then we can omit the hitter of this objective
if it does not appear elsewhere in the graph.
}

\begin{definition}
\label{def:glc}
  The \modMF{saturated} graph of local causality $\cwB \DEF (\Bv, \Be)$ is defined as:
  $\cwB \DEF \lfp{\aB^\w_\ctx}{\myB}{\aB^\w_{\ctx \Cap \allprocs(\myB)}}$,
  where $\myB \DEF (\cwV, \cwE)$ is the smallest graph with
  $\cwV \subseteq \Proc \cup \Obj \cup \Sol$ and $\cwE \subseteq \cwV \times \cwV$
  so that:
  \begin{align}
    \modMF{\toset{\w}} &\subseteq \cwV \label{eq:Vw} \\
%    P \in \cwVObj &\Rightarrow \PHbounce(P) \in \cwV \label{eq:Vproc} \\
    (x, y) \in \cwE &\Rightarrow y \in \cwV \label{eq:VE} \\
    P \in \cwVObj \wedge ps \in \modMF{\aBS}(P) &\Rightarrow (P, ps) \in \cwE \label{eq:ESol1} \\
    ps \in \cwVSol \wedge a_i \in ps &\Rightarrow (ps, a_i) \in \cwE \label{eq:ESol2} \\
    a \in \components \wedge a_i \in \cwVProc \wedge a_j \in \ctx &\Rightarrow (a_i, \PHobjp{a}{j}{i}) \in \cwE \label{eq:EReq} \\
    a \in \cs \wedge a_i \in \cwVProc \wedge ps \in \csState(a_i) &\Rightarrow (a_i, ps) \in \cwE \label{eq:EPrio} \\
    P \in \cwVObj \wedge q \in \gCont_\ctx(\PHsort(P), P) &\Rightarrow (P, \PHobj{q}{\PHbounce(P)}) \in \cwE \label{eq:ECont} \!
  \end{align}
\end{definition}

In the saturated graph of local causality, an edge $(p, ps) \in \Proc \times \Sol$
is said to be coherent (\pref{def:coherent}) if none of the processes in $ps$ conflict with the successor processes of $ps$,
\modMF{
that is, the successors of $ps$ that are in $\Bv \cap \Proc$.
}
Then, \pref{th:approxinf} gives a sufficient condition for the concretisation of a sequence of objectives in a given context,
which is derived immediately from the saturated graph of local causality.
A proof of this theorem is given in \pref{suppl:demoapproxinf}.
\begin{definition}[Coherent edge]
\label{def:coherent}
  An edge $(x, y) \in \cwE$ is said to be coherent if and only if:
  $(x, y) \in \Be \cap (\Proc \times \Sol) \Rightarrow y$ has no successor $a_j \in \Bv \cap \Proc$ so that $\exists a_i \in y$, $a_i \neq a_j$.
\end{definition}

\begin{theorem}[Under-Approximation]
\label{th:approxinf}
Given a Process Hitting $(\PHs; \PHl; \PHa^{\langle 2 \rangle})$
that satisfies \pref{cr:bounded} and \pref{cr:compcs},
a context $\ctx$ and a sequence of objectives $\w$,
  if the graph $\cwB$ contains no cycle,
  all objectives have at least one solution
  and all edges are coherent,
  then $\uconcr(\w) \neq \emptyset$.
\end{theorem}

\modMF{
Regarding the complexity of the method,
computing the saturated graph of local causality is polynomial in the number of sorts in $\PH$ and exponential in the number of processes in one sort.
Checking the properties allowing to apply \pref{th:approxinf} is polynomial in the size of the graph.
Therefore, building and checking process can be considered polynomial in the size
of the PH model, provided that each sort only is contains a few processes.
We note that this is particularly true for biological models, where
each component usually contains a limited number of expression levels.
In addition, it is possible to “factorize” cooperative sorts
modelling more than two components,
as explained in \pref{ssec:cooperation},
in order to reduce their number of processes.
}

\modMF{
We note furthermore that in case the method developed in this section is not conclusive,
it is possible to compute only a subset of $\Bv \cap \Sol$,
by removing some solutions from the initial graph of local causality $\myB$,
or by removing them from the saturated graph of local causality $\cwB$ and by trimming it.
\pref{th:approxinf} is then still valid on the partial graph obtained,
which can lead to more conclusiveness by removing parts of the graph
that were not necessary for the reachability
(such as cycles for instance).
If one wants to test each possible subset of solutions,
the overall method then turns out to be exponential
in the number of solutions to each objective.
Our method can thus be considered as efficient compared to regular model-checking which is usually PSPACE-complete~\cite{Harel02}.
}

\begin{example}
  Let $\PH' = (\PHs, \PHl, \PHh'^{\langle 1 \rangle})$ be the “merged” version of the PH
  depicted in \pref{fig:ph-livelock}, as it was already defined in \pref{ssec:priorities},
  that is: $\PHh'^{\langle 1 \rangle} = \PHh^{(1)} \cup \PHh^{(2)}$.
  In other words,
  $\PH'$ is how $\PH$ would have been represented in the semantics without priorities.
  We showed in \pref{ssec:priorities} that,
  due to spurious behaviours inherent to the cooperative sorts in this semantics,
  process $c_1$ is reachable
  from the initial context $\ctx = \PHstate{a_1, b_0, c_0, ab_{10}}$.
%   This is confirmed by the original under-approximation
%   developed in~\cite{PMR12-MSCS}, which gives the same conclusion.

  Such unwanted behaviours are palliated by the semantics of PH with priorities proposed in this paper.
  Indeed, the under-approximation given in \pref{th:approxinf} does not conclude regarding the reachability of $c_1$,
  as one of the edges of the resulting saturated graph of local causality is not coherent (\pref{def:coherent}),
  as shown in in \pref{fig:sa-livelock}.
  (However, from the inconclusiveness of \pref{th:approxinf}, one cannot conclude about the unreachability of $c_1$.
  Such analysis should be driven for instance with over-approximation methods developed in~\cite{PMR12-MSCS}.)

  We note however that,
  if $\PHhit{a_0}{b_0}{b_1}$ and $\PHhit{b_0}{a_0}{a_1}$ are replaced by the auto-actions
  $\PHhit{a_0}{a_0}{a_1}$ and $\PHhit{b_0}{b_0}{b_1}$,
  then the resulting saturated graph of local causality changes, and
  \pref{th:approxinf} concludes that $c_1$ is then reachable from $\ctx$.

@@

    \node[AprocPrio,below of=c01s] (ab11) {$ab_{11}$};
    \node[AsolPrio,below of=ab11] (ab11s) {};

>>>>

The aim of this section is to define the problem of reachability in an AAN,
and propose an under-approximation allowing to efficiently solve it.
The static analysis presented here is inspired from~\cite{PMR12-MSCS}.
We consider in this section an AAN $\PH = (\PHs; \PHl; \PHa)$.

@@

\modMF{
The main idea behind the static analysis presented in this section
is to abstract the dynamics of an AAN model
by a dynamics that is more general and easier to analyse.
For this, we focus on the notion of objective:
an objective is a couple of local states (\eg $\PHobj{a_i}{a_j}$) whose reachability
is required to ensure the global reachability property considered;
in other words, it is required that a set of actions hitting $a$ exists so that,
starting from a state containing $a_i$,
it is possible to play these actions (possibly intertwined with other actions)
and reach a state containing $a_j$.
Given the particular form of the actions in AANs,
the solving of such an objective raises new objectives in the general case.
%because each action can be triggered by (at most) one local state from a different automaton.
Indeed, each of these local states also have to be reached before the required actions are played.
However, the order of the considered actions is abstracted,
and so is the order of the related objectives,
thus resulting in an over-approximation of the requirements.
}

@@

The reachability of a local state $a_j$ of a given automaton $a$,
starting from another local state $a_i$,
is called an objective and is denoted $\PHobjp{a}{i}{j}$ (\pref{def:obj}).
\begin{definition}[Objective ($\Obj$) \& Objective Sequence ($\OS$)]
\label{def:obj}
  If $a \in \components$, the reachability of a local state $a_j$ from a local state $a_i$ is called an \emph{objective}, noted $\PHobj{a_i}{a_j}$.
  The set of all objectives is called $\Obj \DEF \{ \PHobj{a_i}{a_j} \mid a \in \components \wedge (a_i, a_j) \in \PHl_a \times \PHl_a \}$.
  For an objective $P = \PHobj{a_i}{a_j} \in \Obj$, we define: $\PHsort(P) \DEF
  a$, $\PHtarget(P)\DEF a_i$, $\PHbounce(P)\DEF a_j$,
  and $P$ is said \emph{trivial} if $a_i = a_j$.

  We define an \emph{objective sequence} as a sequence of objectives in which each objective target must be equal to the previous objective bounce of the same automaton, if it exists.
  The set of all objective sequences is denoted by $\OS$.
\end{definition}

A context (\pref{def:context}) extends the notion of state to a set of possible initial states:
\modMF{
to each automaton in the model, a context maps a set of local states in this automaton.
}
We also extend the override operator to contexts (\pref{def:ctxcap}).
\begin{definition}[Context ($\Ctx$)]
\label{def:context}
  A \emph{context} $\ctx$ associates to each automaton in $\PHs$ a non-empty subset of its local states:
  $\forall a \in \PHs, \PHget{\ctx}{a} \subseteq \PHl_a \wedge \PHget{\ctx}{a} \neq \emptyset$.
  $\Ctx$ is the set of all contexts.
\end{definition}
%
\begin{definition}[$\Cap: \Ctx \times \powerset(\PHproc) \rightarrow \Ctx$]
\label{def:ctxcap}
  For any $\ctx\in\Ctx$ and set of local states $ps \in \powerset(\PHproc)$,
  the override of $\ctx$ by $ps$ is noted $\ctx \Cap ps$ and is defined by:
  \[ \forall a \in \PHs, \PHget{(\ctx \Cap ps)}{a} \DEF
  \begin{cases}
    \{ p \in ps \mid \PHsort(p)=a \} & \text{if } \exists p \in ps, \PHsort(p)=a,\\
    \PHget{\ctx}{a} & \text{else.}
  \end{cases}
  \]
\end{definition}
\noindent
For a given context $\ctx$, we note $a_i \in \ctx$ if and only if $a_i \in \PHget{\ctx}{a}$,
and for all $ps \in \powerset(\PHproc)$ or $ps \in \PHl$, $ps \subseteq \ctx \EQDEF \forall a_i \in ps, a_i \in \ctx$.
A sequence of actions $\delta$ is \emph{playable} in a context $\ctx$ if and only if 
$\exists s \subseteq \ctx, \delta \in \Sce(s)$.
We denote then: $\delta \in \Sce(\ctx)$,
and the play of $\delta$ in $\ctx$ is $\ctx \PHplay \delta = \ctx \Cap \ceil(\delta)$,
where $\ceil(\delta)$ is the set containing the last local state in the sequence $\delta$ (hitter or bounce) of every automaton mentioned in $\delta$.

Finally, a bounce sequence on a automaton $a$ (\pref{def:bs}) is a sequence of actions hitting $a$
in which the bounce of each action equals the hitter of the following action.
Bounce sequences are used to find local solutions to a given objective;
\modMF{
indeed, the bounce sequences solving an objective $\PHobj{a_i}{a_j}$
only depend on the bounce part of the actions hitting the automaton $a$
and not on the hitters.
}%
Then, a bounce sequence on $a$ can be abstracted into
its combinations of hitters outside automaton $a$ (\pref{def:aBS});
% the minimal sets of its sets of hitters that are not in automaton $a$ (\pref{def:aBS}).
This abstraction is later used to propagate an objective on a given automaton
by creating new objectives on other automata,
by considering the sets of hitters required by the actions involved.

\begin{definition}[Bounce sequence ($\BS$)]
\label{def:bs}
  A \emph{bounce sequence} $\zeta$ is a sequence of actions so that $\forall n \in \indexes{\zeta}, n < |\zeta|, \PHbounce(\zeta_{n}) = \PHtarget(\zeta_{n+1})$.
  $\BS$ denotes the set of all bounce sequences, and
  $\BS(P)$ denotes the set of bounce sequences \emph{solving} an objective $P$:
  \[
    \BS(\PHobj{a_i}{a_j}) \DEF \{ \zeta \in \BS \mid \PHtarget(\zeta_1)= a_i \wedge \PHbounce(\zeta_{|\zeta|}) = a_j \} \enspace.
  \]
  Furthermore, $\BS(\obj{a_i}{a_j}) = \emptyset$ if there is no way to reach $a_j$ from $a_i$
  and $\emptyseq \in \BS(\obj{a_i}{a_i})$
  \modMF{
  for any trivial objective.
  }
\end{definition}

In the following, we denote: $\sSol = \powerset(\PHproc)$
\moda{%
and $\Sol = \powerset(\sSol)$.
}%
\modMF{%
The elements of $\sSol$ are sets of local states,
and will later represent sets of hitters of a given action
that are required to be active simultaneously in order to play this related action.
The elements in $\Sol$ are sets of sets of hitters,
and represent the requirements to play a whole bounce sequence.
}%

\begin{definition}[$\aBS:\Obj \rightarrow \powerset(\Sol)$]
\label{def:aBS}
  The \emph{abstractions of bounce sequences} of an objective $P$, denoted by the set $\aBS(P)$, are the minimal sets of hitters of the bounce sequences solving $P$:
  \[
    \aBS(P) \DEF \{ \abstr{\zeta} \in \Sol \mid
      \zeta \in \BS(P) \wedge
      \nexists \zeta' \in \BS(P), \abstr{\zeta'} \subsetneq \abstr{\zeta} \} \enspace,
  \]
  \modMF{
  where $\forall P \in \Obj, \forall \zeta \in \BS(P)$:
  $\abstr{\zeta} \DEF \{ \PHhitter(\zeta_n) \mid n \in \indexes{\zeta} \} \setminus \emptyset$.
  }%

  \noindent
  \modMF{
  As a consequence,
  $\aBS(\obj{a_i}{a_j}) = \emptyset$ if $\BS(\obj{a_i}{a_j}) = \emptyset$,
  and $\BS(\obj{a_i}{a_i}) = \{ \emptyset \}$.
  }%
\end{definition}

@@

\noindent
The notation $\delta_{a..b}$ in the previous definition
denotes the subsequence of $\delta$ between indexes $a$ and $b$,
as defined on page \pageref{notations}.

@@

For any objective $P$ and context $\ctx$, \pref{def:maxCont} gives the set of local states of automaton $\PHsort(P)$ that are required to solve $P$ in $\ctx$, denoted by $\gCont_\ctx(\PHsort(P), P)$.
\begin{definition}[$\gCont_\ctx : \Sigma \times \Obj \rightarrow \powerset(\PHproc)$]
  \label{def:maxCont}
  \begin{align*}
    \gCont_\ctx(a,P) \DEF
    \{ p \in \PHproc &\mid \exists ps \in \aBS(P), \exists b_i \in ps, b = a \wedge p = b_i \\
      & \vee b \neq a \wedge p \in \gCont_\ctx(a, \PHobj{b_j}{b_i}) \wedge b_j \in \PHget{\ctx}{b} \}
    \enspace.
  \end{align*}
\end{definition}

\modMF{
A graph of local causality $\myB = (\cwV, \cwE)$
is a graph where $\cwV \subseteq \Proc \cup \Obj \cup \sSol \cup \Sol$ is the set of vertices
and $\cwE \subseteq \cwV \times \cwV$ is the set of oriented edges.
Such a graph aims a describing the requirements of a given reachability
problem, that is, the reachability of $\w$ from the initial context $\ctx$.
Thus, a node in $\PHproc$ represents a local state required to play an action,
a node in $\Obj$ is an objective to reach a given local state,
a node in $\sSol$ is a set of hitters that have to be active simultaneously
to solve one step of a given objective,
and a node in $\Sol$ is a set of sets of hitters,
thus encompassing all requirement for the solving of a whole objective.
}

\modMF{
The edges of such a graph allow to draw links between these requirements.
Thus, an objective $P \in \Obj$ is solvable if
at least one related bounce sequence in $\BS(P)$ can be played;
therefore, we can consider all minimal sets of sets of hitters of these bounce sequences
(that is, the sets of $\aBS(P) \subseteq \Sol$, see \pref{def:aBS})
as requirements to solve this objective (\pref{eq:ESol1}).
Such a set can then naturally be split into as many sets of hitters (\pref{eq:ESol3})
which all have to be simultaneously active.
These sets of hitters can in turn be split into several local state nodes (\pref{eq:ESol2}).
This creates new requirements, as the reachability of a required local state $a_i$
is approximated by the ability to solve all objectives of the form
$\PHobjp{a}{j}{i} \in \Obj$ for all $a_j$ in the initial context (\pref{eq:EReq}).
Furthermore, we note that the solving of an objective $P$ may require
a local state of $\PHsort(P)$,
\ie $\gCont(\PHsort(P), P) \neq \emptyset$ (see \pref{def:maxCont});
in this case, $P$ is re-targeted (\pref{eq:ECont}).
Finally, \pref{eq:Vw} forces the graph to contain all the initial objectives,
and \pref{eq:VE} ensures the consistency of $\cwV$.
}

\modMF{
However, this graph $\myB$ alone is not sufficient to check the reachability of
an objective sequence $\w$ from an initial context $\ctx$;
indeed, the reachability of each intermediate required local state
is only checked starting from the context $\ctx$,
although this context may be modified by other objectives.
Therefore, such a reachability requires the building of the
\emph{saturated graph of local causality} (\pref{def:glc})
which is obtained by iteratively saturating the initial context with all
local states referenced in the graph,
until reaching a fixed point.
Such a fixed point always exists because the size of the context
is limited by the size of the initial PH model.
The set of local states contained in a graph of local causality
is given by:
\begin{align*}
  \allprocs(\cwV, \cwE) = (\cwV \cap \Proc) &\cup \{ \PHtarget(P) \mid P \in \cwV \cap \Obj \}\\
  & \cup \{ P \neq \w \Rightarrow \PHbounce(P) \mid P \in \cwV \cap \Obj \}
\end{align*}
where, if the objective sequence $\w$ consists of only one element,
we can omit the hitter of this objective
provided that it does not appear elsewhere in the graph.
}

\begin{definition}
\label{def:glc}
  The \emph{\modMF{saturated} graph of local causality} $\cwB \DEF (\Bv, \Be)$ is defined as:
  $\cwB \DEF \lfp{\aB^\w_\ctx}{\myB}{\aB^\w_{\ctx \Cap \allprocs(\myB)}}$,
  where $\myB \DEF (\cwV, \cwE)$ is the smallest graph with
  $\cwV \subseteq \Proc \cup \Obj \cup \sSol \cup \Sol$ and $\cwE \subseteq \cwV \times \cwV$
  so that:
  \begin{align}
    \modMF{\toset{\w}} &\subseteq \cwV \label{eq:Vw} \\
%    P \in \cwVObj &\Rightarrow \PHbounce(P) \in \cwV \label{eq:Vproc} \\
    (x, y) \in \cwE &\Rightarrow y \in \cwV \label{eq:VE} \\
    P \in \cwVObj \wedge pps \in \modMF{\aBS}(P) &\Rightarrow (P, pps) \in \cwE \label{eq:ESol1} \\
    \moda{pps \in \cwVSol \wedge A \in pps} &\moda{\Rightarrow (pps, A) \in \cwE} \label{eq:ESol3} \\
    A \in \cwVsSol \wedge a_i \in A &\Rightarrow (A, a_i) \in \cwE \label{eq:ESol2} \\
    a_i \in \cwVProc \wedge a_j \in \ctx &\Rightarrow (a_i, \PHobjp{a}{j}{i}) \in \cwE \label{eq:EReq} \\
%    a \in \cs \wedge a_i \in \cwVProc \wedge ps \in \csState(a_i) &\Rightarrow (a_i, ps) \in \cwE \label{eq:EPrio} \\
    P \in \cwVObj \wedge q \in \gCont_\ctx(\PHsort(P), P) &\Rightarrow (P, \PHobj{q}{\PHbounce(P)}) \in \cwE \label{eq:ECont} \!
  \end{align}
\end{definition}

\modMF{
In the saturated graph of local causality, a set of hitters $A \in \sSol$
is said to be \emph{coherent} (\pref{def:coherent})
if none of the local states in $A$ conflict with
a successor of $A$ in $\Bv \cap \Proc$,
that is, if there exists no local state node $a_j$ in the successors of $A$
and no element $a_i \in A$ so that $a_i \neq a_j$.
}
Then, \pref{th:approxinf} gives a sufficient condition for the concretisation
of an objective sequence in a given context,
which is derived immediately from the saturated graph of local causality,
and the condition that all sets of hitters are coherent.
A proof of this theorem is given in \pref{suppl:demoapproxinf}.
\begin{definition}[Coherent node]
\label{def:coherent}
  \modMF{
  In a given saturated graph of local causality $\cwB = (\Bv, \Be)$,
  a node $x \in \Bv$ is said to be \emph{coherent} if and only if:
  $x \in \Bv \cap \sSol \Rightarrow x$ has no successor $a_j \in \Bv \cap \Proc$
  so that $\exists a_i \in x$, $a_i \neq a_j$.
  }
\end{definition}

\begin{theorem}[Under-Approximation]
\label{th:approxinf}
  Given an AAN $(\PHs; \PHl; \PHa)$,
  a context $\ctx$ and an objective sequence $\w$,
  if the graph $\cwB$ contains no cycle,
  all objectives have at least one solution
  and all nodes are coherent,
  then $\uconcr(\w) \neq \emptyset$.
\end{theorem}

\modMF{
Regarding the complexity of the method,
computing the saturated graph of local causality is polynomial in the number of automata in $\PH$ and exponential in the number of local states in one automaton.
Checking the properties allowing to apply \pref{th:approxinf} is polynomial in the size of the graph.
Therefore, the building and checking process can be considered as polynomial in the size
of the AAN, provided that each automaton only contains a few local states.
We note that this is particularly true for biological models, where
each component usually contains a limited number of expression levels.
}

\modMF{
We note furthermore that in the case where
the method developed in this section is not conclusive,
it is possible to compute only a subset of $\Bv \cap \Sol$,
by removing some solutions from the initial graph of local causality $\myB$,
or by removing them from the saturated graph of local causality $\cwB$ and by trimming it.
In other words, this consists in ignoring some of the bounce sequences
for some of the objectives,
which intuitively cannot create false positives.
Indeed, \pref{th:approxinf} is then still valid on the partial graph obtained,
and this removal can lead to more conclusiveness by trimming parts of the graph
that were not necessary for the reachability
(especially cycles or unnecessary but inconclusive branches).
If one wants to try each possible subset of solutions,
then the overall method turns out to be exponential
in the number of solutions to each objective.
Our method can thus still be considered as efficient
compared to regular model-checking which is usually PSPACE-complete~\cite{Harel02}.
}

\begin{example}
  \moda{%
  If we confider the AAN of \pref{fig:ph-livelock},
  from the initial state $\PHstate{a_1, b_0, c_0}$ depicted,
  the under-approximation given in \pref{th:approxinf}
  does not conclude regarding the reachability of $c_1$.
  This is due to the fact that the node $\{ a_1, b_1 \} \in \Bv \cap \sSol$
  is not coherent because of its successor $a_0$ (and $b_0$).
  (However, from the inconclusiveness of \pref{th:approxinf},
  one cannot conclude about the unreachability of $c_1$.
  Such analysis should be driven for instance
  with over-approximation methods developed in~\cite{PMR12-MSCS}.)
  }%

  \moda{%
  This result is new compared to the method proposed in~\cite{PMR12-MSCS}.
  Indeed, the representation based on the Process Hitting that was proposed
  in this paper only allowed to represent “over-approximated” Boolean gates
  with the use of cooperative sorts,
  as explained in \pref{ssec:cooperation}.
  This especially did not allow to model the fact that $a_1$ and $b_1$ could not
  be activated in the same state, but only in successive states.
  Thus, $c_1$ was indeed reachable,
  contrary to the behaviour expected from an accurate Boolean gate.
  }%

  \moda{%
  Finally, we note however that,
  if $\PHhits{a_0}{b_0}{b_1}$ and $\PHhits{b_0}{a_0}{a_1}$ are replaced by the actions
  $\PHhitm{a_0}{a_1}$ and $\PHhitm{b_0}{b_1}$,
  then the resulting saturated graph of local causality changes, and
  \pref{th:approxinf} concludes that $c_1$ is reachable from $\ctx$.
  }%

@@

    \node[Assol,below of=c01s] (a1a1ss) {$\{ a_1, b_1 \}$};
    \node[Aproc,below left of=a1a1ss] (a1) {$a_1$};
    \node[Aobj,below of=a1] (a11) {$\PHobj{a_1}{a_1}$};
    \node[Asol,below of=a11] (a11s) {};
    \node[Assol,below of=a11s] (na11s) {$\emptyset$};
    \node[Aobj,below left of=a1] (a01) {$\PHobj{a_0}{a_1}$};
    \node[Asol,below of=a01] (a01s) {};
    \node[Aproc,below of=a01s] (b0) {$b_0$};
    \node[Aobj,below of=b0] (b00) {$\PHobj{b_0}{b_0}$};
    \node[Asol,below of=b00] (b00s) {};
    \node[Assol,below of=b00s] (nb00s) {$\emptyset$};
    \node[Aobj,below left of=b0] (b10) {$\PHobj{b_1}{b_0}$};
    \node[Asol,below of=b10] (b10s) {};
    \node[Assol,below of=b10s] (nb10s) {$\emptyset$};
